From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-07 Pi+/paper/discussion.tex
intro

@@

>>>>

Pi types -- Natural number -- Finite sets
1-combinator -- Generators of Sn -- 1-paths
2-combinator -- Relations of Sn -- 2-paths

@@

\paragraph{Extensional vs. Intensional View of Permutations.}
Although the meaning of a permutation - that is, the specific bijective function
that it represents - is in some sense all there is, manipulating them
syntactically still has its advantages. By writing reversible programs, we think
of them in an intensional way. Comparing two programs for equality by evaluating
them on all points in the domain is a very crude - and maybe even inefficient -
way of doing that. ~\cite{Kuehlmann:2006:RBR:2298470.2300327,10.1007/978-3-540-24605-3_4,Yamashita:2010:FEQ:1835957.1835965}.

It is crude, because there is no way of enforcing additional constraints on the
process of transforming one program into another, and no way of inspecting what
transformation occurred. We can imagine a practical situation in which reversible
circuits admit one kind of optimization FPGA?, but do not admit another,
even though they are equivalent in the extensional sense - or maybe one of these
transformations is cheaper than another, or maybe it is important to know which
transformation did occur for the producer(?) to focus their resources of
improving this kind of transformations. By comparing the circuits extensionally,
the proof object = the path from one into the other - does not have any
structure,  it is just a check that the values match.

On the other hand, even though in general, the equality of circuits requires
exponential time jk{triple check}, the proof object can be still very small.
This creates structure that can possibly be exploited in a heuristic way.
Additionally, if the goal is to convince third-party that two circuits are
the same, we can just present the proof instead of showing the equality directly
- for example, by choosing a particularly short one.

Taking all this into account, it is clear that we must seek a syntactic way of
computing the meaning of the circuit, instead of just directly evaluating it to
$\Aut[\Fin[n]]$ - since when we go back and quote the permutation as a circuit
(in a normal form), there was no "trace" left to see what transformation (what
sequence of 2-paths) maps the old one into the new one.

We are additionally experimenting with user annotations that can guide
the search. Each level-2 combinator can be annotated with various
``cost'' annotations indicating whether it reduces the number of
gates, reduces the number of choice points, or other cost
functions. Then one can ask for a proof that takes no more than a
certain number of steps or a proof that does not create more than a
certain number of additional wires etc. We illustrate these ideas by
defining a simple cost function and using it to annotate level-2
combinators.

We define the \emph{length} $L(c)$ of a composite circuit $c$ as
follows: the length of a sequential composition of circuits is the sum
of the lengths of the subcircuits $L(f \odot g) = L(f) + L(g)$; and
the length of choice or parallel composition is the maximum of either
branch $L(f \oplus g) = L(f \otimes g) = \max(L(f),L(g))$. For
primitive gates, the length needs to be postulated to reflect the
``length'' of the computation involved in applying that primitive. As
examples, consider the following two level-2 combinators:

% \begin{code}
% linv◎l'  :  {t₁ t₂ : U} {c : t₁ ⟷ t₂} → (c ◎ ! c) ⇔ id⟷
% idl◎l'   :  {t₁ t₂ : U} {c : t₁ ⟷ t₂} → (id⟷ ◎ c) ⇔ c
% \end{code}
% \AgdaHide{
% \begin{code}
% linv◎l'  = ?
% idl◎l'   = ?
% \end{code}
% }

\noindent Assuming that $\idc$ takes a unit length of computation, the
first can be annotated with $L(c)*2 \isotwo 1$ and the second with $L(c)+1 \isotwo
L(c)$ indicating that the first combinator reduces the length of the
circuit from twice the length of $c$ to 1 and the second combinator
reduces the length of the circuit by 1. Such annotations can then be
used to constrain or guide the search for transformations between
circuits.
